TAPIR: Threaded Application Programming on Intermittent Resources

### 2017-05-11 paper.tex
Added section on scheduler data structures Added style to make code listings nicer Trailing whitespace fixes

\usepackage{balance}
\usepackage{comment}
\usepackage{listings}
\usepackage{xspace}
\usepackage[export]{adjustbox}
\usepackage{wrapfig}
\newcommand{\chain}{Chain\xspace}
\numberofauthors{3}

@@

@@

% Doesn't compile, commenting out
\begin{figure}[h]
  \centering
  \includegraphics[width=0.99\columnwidth]{figs/listing}
  \caption{{\bf Example program for sensing and sending data}}
  \label{fig:copy}
\end{figure}

@@

\section{System Implementation } % TODO change title
Mutexes - Emily\\

TAPIR has a round robin scheduling algorithm. At the end of a task, the compiler
inserts a jump back to the scheduler task, where the next task in the list is
the one to be executed. An alternate scheduling model that was considered was
a context switch on reboot instead of at the end of each task. The main drawback
of using a reboot-based scheduler is that, because power failure is the common
case, there is a high chance of reducing or halting forward progress altogether
because one thread holds a lock that all of the other threads need but cannot
complete. For that reason the design specifies end of task, so each task has
an opportunity to use a shared resource and release any locks before a power-
intensive context switch occurs. Additionally, a round robin scheduler was 
chosen for this design because it is simple to implement and uses only one index counter,
very little power or memory overhead as compared to other scheduling algorithms. It is 
critical for multithreading in Chain to not add significant power and memory
overhead in scheduling over conventional Chain so that the workload can
make as much forward progress as possible on one charge cycle.

When designing threading in Chain, the shared memory model is a main indicator 
of the threaded execution model. The two most probable options when designing 
TAPIR were a SIMD execution model of \textit{Thread Pools}, or a MIMD 
traditional multiprogramming model. The Thread Pool design makes use of task 
code duplication by explicitly forking a number of chain programs, all of which 
run simultaneously on the scheduler on arrayed data. This option is useful 
multiple data streams from I/O in embedded applications that need to be 
processed before new data overrides the I/O device. Additionally, the Thread 
Pool model makes sharing memory more regular by using multiple copies of the 
same thread. However, this option is not as general purpose. The TAPIR design 
ultimately settled on a multiprogramming model, because this allows for the 
programmer to explictly program unique chain programs with shared or discrete 
data structures on the same system. However, this execution model requires a
more structured locking mechanism to avoid RMW during execution.

SIMD w/ channels (you don't get your own registers...etc -
    just channels, since intermittent)\\

@@

\subsection{Locking Primitive} 
To improve a programmer's control of shared data between threads, and to reduce
the complexity of accessing data that may be touched by many different tasks we
implement a simple mutex object that can be added to a channel. The mutex is
designed to be used in a standard \texttt{channel}, not a double buffered
\texttt{self\_channel}, so it must be able to tolerate a failure after a
read/write operation (i.e. test and set). To achieve idempotence, we add a
\texttt{thread\_id} to the \texttt{free} bit inside of each mutex. When
attempting to acquire a lock, a thread checks if its thread id is equal to the
lock holder, then it checks if the lock is free. If either condition is true,
the thread successfully acquires the lock, otherwise the \texttt{mutex\_lock}
function fails. Upon successfully acquiring the lock, a thread writes its id to
the mutex's holder id, and changes the \texttt{free} bit to \texttt{status =
held}. This scheme guarantees correctness because if a failure occurs after the
mutex is set but before the task commits, the task will read out the id of the
mutex holder before checking if it is free. Upon matchint its id with the holder
id, the task will proceed regardless of the fact that \texttt{status = held}.
%Maybe add a figure here to make the sequence of read/write/failure clearer... 
Further, a task must finsih executing before switching to another thread, so
there is no possiblity for an external thread to see a value that was not
committed by another task.

It is important to note that the mutex object can be modified idempotently, but
accesses to the data in the same channel as the mutex must adhere to channel
exclusion. In our example application, the mutex is contained within a user
defined \texttt{shared\_channel} structure that is analogous to a
\texttt{multicast\_channel} but abandons the naming convention that "restricts"
the channel to a static subset of tasks. Any task may read {\em or} or write to
the \texttt{shared\_channel}, but performing both would break correctness. In
the future, a compiler could be used to guarantee that a
\texttt{shared\_channel} is not read and written by the same task, but at
present we believe that the \texttt{mutex} and \texttt{shared\_channel}
interface is easier for programmers to use than \texttt{multicast\_channels}, or
a sepcialized double buffered channel when passing data between threads. It
maintains familiar semantics for programmers and does not incur the memory
overhead of double buffering.

@@

\begin{figure*}
\begin{minipage}[b]{1.0\textwidth}
\begin{lstlisting}
    /** Enable interrupts if int_setup_complete() has been called
      * and execution is not inside an interrupt handler */
    void enable_interrupts();

@@

\section{Implementation}
Scheduler Data structures and idempotence (what we broke and why) -NEIL\\
Mutexes (Simple, meant for pseudo channels) - Emily\\
Interrupts (mucked with IE flags) - Neil\\
% Interrupt prologue
% Interrupt setup

>>>>

\usepackage{balance}
\usepackage{comment}
\usepackage{listings}
\usepackage{xspace}
\usepackage[export]{adjustbox}
\usepackage{wrapfig}
\newcommand{\chain}{Chain\xspace}

\usepackage{color}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.95}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=none,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=4
    }
    \lstset{style=mystyle}

\numberofauthors{3}

@@

%Example Citation\cite{Chain, Mementos, Aware, Dino}!

%Sample of adding a picture across the top of a split page...

%\begin{figure*}
%\centering
%\begin{minipage}[b]{0.49\textwidth}
%  \includegraphics[width=0.95\textwidth,center]{capybara-all.png}
%\caption{Capybara power system overview}\label{label-a}
%\end{minipage}\hfill
%\begin{minipage}[b]{0.49\textwidth}
%  \includegraphics[width=0.95\textwidth,center]{power_modes.pdf}
%\caption{Effect of different power modes}\label{label-b}
%\end{minipage}
%\end{figure*}

%More text after the picture to show how that looks...

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.99\columnwidth]{figs/listing}
  \caption{{\bf Example program for sensing and sending data}}
  \label{fig:copy}
\end{figure}

@@

\section{What we did} % TODO change title
Mutexes - Emily\\
Round Robin (alternatives, on reboot)\\
SIMD vs MIMD\\
SIMD w/ channels (you don't get your own registers...etc -
    just channels, since intermittent)\\

@@

@@

\begin{figure*}
\begin{minipage}[b]{1.0\textwidth}
\begin{lstlisting}[language=C]
    /** Enable interrupts if int_setup_complete() has been called
      * and execution is not inside an interrupt handler */
    void enable_interrupts();

@@

\section{Implementation}
\subsection{Scheduler}
\begin{figure*}
\begin{minipage}[b]{.48\textwidth}
\begin{lstlisting}[language=C]
    typedef struct thread_t {
        unsigned thread_id;
        context_t context;
    } thread_t;

    typedef struct thread_state_t {
        thread_t thread;
        unsigned active;
    } thread_state_t;

    // Array of running threads
    thread_state_t threads[MAX_NUM_THREADS];

    // List of free indicies in threads[]
    unsigned free_indicies[MAX_NUM_THREADS];
    unsigned free_indicies_size;

    // Index of current running thread in threads[]
    // Stored in self channel
    unsigned current;
\end{lstlisting}
\caption{Scheduler Data Structures}\label{schDataStruct}
\end{minipage}\hfill
\end{figure*}

The data structures for our scheduler are listed in figure \ref{schDataStruct}.
Like the rest of chain, memory for data structures is allocated at compile
time. Thus, the limit on the maximum number of threads allowed in a system is
fixed arbitrarily - not ideal, but there isn't a better option while operating
inside chain. The \texttt{threads} array is stored in all-to-one channel, from
every task to the scheduler task. The \texttt{free\_indicies} array is stored
in a one-to-all channel, from the scheduler task to every task.

Neither all-to-one nor one-to-all channels fit into standard \chain channel
semantics, although they resemble multicast-in and multicast-out channels
described in the original \chain paper. We eschew the notion that channels
have fixed start and end points and instead treat the channel as a general
location in non-volatile memory that can be read from and written to by
arbitrary tasks. However, accesses to these channels are encapsulated by our
library, so the programmer has no need to think about idempotency violations.

With these multicast channels, the scheduler task can act as a buffer between
multiple threads - arbitrary tasks write \texttt{threads[]} with calls to
\texttt{thread\_create} and \texttt{thread\_end} and tasks read from
\texttt{free\_indicies} through calls to thread create, maintaining
idempotency. Inside the scheduler task, we read and write \texttt{current},
so \texttt{current} is double buffered. The only issue is with
\texttt{thread\_init} since it is called from user-defined tasks and
initializes \texttt{current} - the same task that called \texttt{thread\_init}
could call \texttt{thread\_create}, and the \texttt{thread\_create} call
wouldn't see the update to the variable. Since \texttt{current} is always
initialized to the same value (0), we opt for manually swapping the buffer on
\texttt{current} so that other threads can see the update.

We keep an unsigned integer, \texttt{curr\_free\_index}, in volatile memory,
zeroed on every reboot, that acts as a current pointer in
\texttt{free\_indicies[]}. Every time the scheduler task runs,
\texttt{free\_indicies[]} is updated with the state of \texttt{threads[]},
based on whether or not the active bit is set for a given index. When
\texttt{thread\_create} is called, if \texttt{curr\_free\_index} is greater
than \texttt{free\_indicies\_size}, then \texttt{thread\_create} will fail in
POSIX fashion - the system does not have enough resources to support another
thread. Otherwise, \texttt{threads[free\_indicies[curr\_free\_index]]} is set
to the new thread. Upon power failure, \texttt{curr\_free\_index} is reset to
zero, so as long as the rest of the operations in the task are idempotent, so
will \texttt{thread\_create} calls - the same locations in \texttt{theads[]}
will be overwritten.

Mutexes (Simple, meant for pseudo channels) - Emily\\
Interrupts (mucked with IE flags) - Neil\\
% Interrupt prologue
% Interrupt setup
